Parallel Adaptive Importance Sampling

### 2016-06-02 PAIS.tex
Some changes
Not very informative really!

 Using an adaptive strategy, we calculate a sequence
 $\{\beta^{(k)}\}_{k=1}$ which converges roughly to the optimal
 scaling parameter $\beta^*$, resulting in the optimal transition
 density $\chi$ for our MCMC algorithm. We must choose some sequence
 of iterations, $\{n_k\}_{k=1}$, at which to update $\beta$, and due
 to the constraints on adaptive MCMC
 algorithms\cite{roberts2007coupling,roberts2009example}, these $n_k$
 must grow exponentially further apart. This same adaptive approach
 can also be applied to the trivially parallelised MCMC algorithms to adaptively calculate their optimal scaling parameter $\beta^*$.

@@

Since we are comparing against naively parallelised MH algorithms,
we also need to decide which statistic $T(\beta)$ to optimise for
these approaches. In the examples which follow, we have optimised the naively parallelised RWMH algorithm using the optimal acceptance rate $\hat{\alpha} = 0.5$. This value differs from the theoretical asymptotic value of 0.234 which applies in higher dimensions, this higher acceptance rate is commonly used for one dimensional Gaussian posteriors~\cite{rosenthal2011optimal}. To optimise the algorithm we minimise the statistic
\[
	T_{\text{MH}}(\beta) = \left| \frac{N_{\text{acc}}(\beta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\beta)$ is the number of accepted moves and
$N_{\text{total}}$ is the total number of samples produced. For the
PAIS algorithm, we are maximising the effective sample size as
discussed in Section~\ref{sec:ess}, so $T_{\text{PAIS}}(\beta) =
\neff(\beta)$.

@@

Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow Gaussian distributions. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|u\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
we observe $u=4$ noisily such that
\[
D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\sigma^2).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this problem.

@@

Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio compared to the error analysis and the variance of the weights. The relative $L^2$ error graph is noisy, but it is clear that the maximum in the effective sample size and the minimum in the variance of the weights are both close to the minimum in the relative $L^2$ error. Due to this we say that the estimate of the effective sample size found by averaging the statistic over each iteration is a good indicator for the optimal scaling parameter. In general this indicator underestimates the estimate found by using the entire sample.

@@

Both adaptive algorithms are run with initial values of $\beta_0=2.0$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance. The ARWMH algorithm does not perform quite as well, this is possibly due to the fact that the acceptance rate cost function is not smooth at the optimal value making it difficult to minimise.

@@

Throughout this paper, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales when we increase the ensemble size, and if there is some limit below which the algorithm fails. We implement the problem in Section~\ref{sec:problem 1}, using the RWMH and PAIS-RW algorithms with ensemble sizes in the inteveral $M \in [1, 160]$.

@@

Figure~\ref{fig:PAIS_saving} was produced using the method of finding
optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
then running 32 repeats at each ensemble size. The convergence rates
are then found by regressing through the data. The graph is still very
noisy but demonstrates that increasing the ensemble size continues to
reduce the number of iterations required in comparison with naively parallelised MH. When $M<8$, the algorithm takes a long time to reach
stationarity. The decreaseing trend indicates superlinear improvement of PAIS with
respect to ensemble size, in terms of the number of iterations
required, which is a demonstration of our belief that parallelism of
MCMC should give us added value over and above that provided by naive parallelism. This decrease is linked to the increasing effective sample size shown in Figure~\ref{fig:neff-M} (b).

@@

We look at an `easy' problem, $B_1$, which has a KL divergence of 0.880, and a `harder' problem, $B_2$, which has a KL divergence of 3.647. Problem $B_1$ has two modes which are separated by a small energy barrier. In $B_2$ we increase the distance between the two modes which has the effect of increasing the required energy to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.

\subsubsection{Target Distribution}

@@

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{"figures/posteriors3"}
\caption{Posterior distributions for problem $B_1$ and $B_2$. Problem $B_1$ has a noisy data value of  0.921312 which results in an energy barrier which is relatively easy to cross, as well as their common prior distribution.}
\label{fig:problem 3 posteriors}
\end{center}
\end{figure}

@@

Calculating the optimal values of the scaling parameters for this problem is similar to the Gaussian case; we check only the acceptance rate to find the optimal values for RWMH and we use the effective sample size to find the optimal values for PAIS-RW. Table~\ref{table:BM_opt_beta} gives the optimal values of $\beta$ for both problems.

@@

It is relatively simple to find optimal scaling parameters for problem $B_1$. These values are given in Table~\ref{table:BM_opt_beta} (left). However problem $B_2$ is much harder; transitions between the modes are extremely unlikely for the standard RWMH algorithm. This means that we need to consider the
convergence on two levels; we should consider the algorithm's ability
to find all the modes, and also whether it can sample them in the correct proportions.

@@

From Table~\ref{table:BM_opt_beta} (right) we see that there is a large difference between the optimal value of $\beta$ for each regime in RWMH, and so will result in inefficient sampling. The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\beta^*$: a clear advantage to using this algorithm for this problem.

@@

As in the Gaussian example we see a significant speed up with the PAIS-RW algorithm for $B_1$. Figure~\ref{fig:BM1_L2} shows the adaptive and nonadaptive convergence rates. We can see that the adaptive algorithms compare closely with the
respective nonadaptive algorithms and the improvement PAIS offers remains significant.

@@

Comparing the convergence of the adaptive algorithms against the nonadaptive algorithms in Figure~\ref{fig:BM2_AL2} shows that the algorithms behave as expected. The adaptive RWMH algorithm tuned using the acceptance rate converges at the same rate as the $L^2$ optimised algorithm until the scaling parameter gets small where the algorithm forgets one of the modes. The adaptive PAIS-RW with scout chains tuned to the effective sample size converges at about the same rate as the locally optimised nonadaptive algorithm also with scouts.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{"figures/calc_saving"}
\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $10^{-2.5}$ as a percentage of MH iterations. The relative $L^2$ error graphs are from the Gaussian problem in Section~\ref{sec:problem 1}.}
\label{fig:calc_saving}
\end{center}
\end{figure}

>>>>

 Using an adaptive strategy, we calculate a sequence
 $\{\beta^{(k)}\}_{k=1}$ which converges roughly to the optimal
 scaling parameter $\beta^*$, resulting in the optimal transition
 density $\chi$ for our MCMC algorithm. We must choose some sequence
 of iterations, $\{n_k\}_{k=1}$, at which to update $\beta$, and due
 to the constraints on adaptive MCMC
 algorithms\cite{roberts2007coupling,roberts2009examples}, these $n_k$
 must grow exponentially further apart. This same adaptive approach
 can also be applied to the trivially parallelised MCMC algorithms to adaptively calculate their optimal scaling parameter $\beta^*$.

@@

Since we are comparing against naively parallelised MH algorithms,
we also need to decide which statistic $T(\beta)$ to optimise for
these approaches. In the examples which follow, we have optimised the
naively parallelised RWMH algorithm using the optimal acceptance rate
$\hat{\alpha} = 0.5$. This value differs from the theoretical
asymptotic value of 0.234 which applies in higher dimensions, but this higher acceptance rate is commonly used for one dimensional Gaussian posteriors~\cite{rosenthal2011optimal}. To optimise the algorithm we minimise the statistic
\[
	T_{\text{MH}}(\beta) = \left| \frac{N_{\text{acc}}(\beta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\beta)$ is the number of accepted moves and
$N_{\text{total}}$ is the total number of samples produced. For the
PAIS algorithm, we are maximising the effective sample size as
discussed in Section~\ref{sec:ess}.

@@

Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow Gaussian distributions. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|u\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
we observe $u=4$ noisily such that
\[
D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\sigma^2) = \mathcal{N}(4,0.01).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this
problem. A KL divergence of zero indicates that two distributions are
identical almost everywhere.

@@

Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio compared to the error analysis and the variance of the weights. The relative $L^2$ error graph is noisy, but it is clear that the maximum in the effective sample size and the minimum in the variance of the weights are both close to the minimum in the relative $L^2$ error. Due to this we say that the estimate of the effective sample size found by averaging the statistic over each iteration is a good indicator for the optimal scaling parameter. In general this indicator underestimates the value found by using the entire sample.

@@

Both adaptive algorithms are run with initial values of $\beta=2.0$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance. The ARWMH algorithm does not perform quite as well, this is possibly due to the fact that the acceptance rate cost function is not smooth at the optimal value making it difficult to minimise.

@@

Throughout this example, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales when we increase the ensemble size, and if there is some limit below which the algorithm fails. We implement the problem in Section~\ref{sec:problem 1}, using the RWMH and PAIS-RW algorithms with ensemble sizes in the inteveral $M \in [1, 160]$.

@@

Figure~\ref{fig:PAIS_saving} was produced using the method of finding
optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
then running 32 repeats at each ensemble size. The convergence rates
are then found by regressing through the data. The graph is still very
noisy but demonstrates that increasing the ensemble size continues to
reduce the number of iterations required in comparison with naively parallelised MH. When $M<8$, the algorithm takes a long time to reach
stationarity. The decreasing trend indicates superlinear improvement of PAIS with
respect to ensemble size, in terms of the number of iterations
required, which is a demonstration of our belief that parallelism of
MCMC should give us added value over and above that provided by naive parallelism. This decrease is linked to the increasing effective sample size shown in Figure~\ref{fig:neff-M} (b).

@@

We look at an `easy' problem, $B_1$, which has a KL divergence of
0.880, and a `harder' problem, $B_2$, which has a KL divergence of
3.647. Problem $B_1$ has two modes which are not too far apart. In
$B_2$ we increase the distance between the two modes which has the
effect of increasing the expected number of iterations that it takes
for a MCMC chain to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.

\subsubsection{Target Distribution} \label{sec:tar}

@@

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{"figures/posteriors3"}
\caption{Posterior distributions for problems $B_1$ and $B_2$ as
  described in Section \ref{sec:tar}.}
\label{fig:problem 3 posteriors}
\end{center}
\end{figure}

@@

Calculating the optimal values of the scaling parameters for this
problem is similar to the previous example; we check only the acceptance rate to find the optimal values for RWMH and we use the effective sample size to find the optimal values for PAIS-RW. Table~\ref{table:BM_opt_beta} gives the optimal values of $\beta$ for both problems.

@@

It is relatively simple to find optimal scaling parameters for problem $B_1$. These values are given in Table~\ref{table:BM_opt_beta} (left). However problem $B_2$ is much harderas transitions between the modes are extremely unlikely for the standard RWMH algorithm. This means that we need to consider the
convergence on two levels; we should consider the algorithm's ability
to find all the modes, and also whether it can sample them in the correct proportions.

@@

From Table~\ref{table:BM_opt_beta} (right) we see that there is a large difference between the optimal value of $\beta$ for each regime in RWMH, and so will result in inefficient sampling. The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\beta^*$; a clear advantage to using this algorithm for this problem.

@@

As in the Gaussian example we see a significant speed up with the
PAIS-RW algorithm for problem $B_1$. Figure~\ref{fig:BM1_L2} shows the adaptive and nonadaptive convergence rates. We can see that the adaptive algorithms compare closely with the
respective nonadaptive algorithms and the improvement PAIS offers remains significant.

@@

Comparing the convergence of the adaptive algorithms against the
nonadaptive algorithms in Figure~\ref{fig:BM2_AL2} shows that the
algorithms behave as expected. The adaptive RWMH algorithm tuned using
the acceptance rate converges at the same rate as the $L^2$ optimised
algorithm until the scaling parameter gets small, and therefore
switches between the modes are rare, and the relative heights of the
modes are decided by the arbitrary proportion of chains which are in
each mode at this point. The adaptive PAIS-RW with scout chains tuned to the effective sample size converges at about the same rate as the locally optimised nonadaptive algorithm also with scouts.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{"figures/calc_saving"}
\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $~10^{-2.85}$ as a percentage of MH iterations. The relative $L^2$ error graphs are from the Gaussian problem in Section~\ref{sec:problem 1}.}
\label{fig:calc_saving}
\end{center}
\end{figure}
